### 2020-01-27 Mathematical Statistics/05_conf_int_bootstrap.rmd
bootstrapping
bootstrapping

library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(ggthemes, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
library(Rblpapi, quietly = TRUE, warn.conflicts = FALSE)
library(scales, quietly = TRUE, warn.conflicts = FALSE)
library(pander, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(formattable, quietly = TRUE, warn.conflicts = FALSE)
library(grid, quietly = TRUE, warn.conflicts = FALSE)
library(gridExtra, quietly = TRUE, warn.conflicts = FALSE)
library(png, quietly = TRUE, warn.conflicts = FALSE)
library(extrafont, quietly = TRUE, warn.conflicts = FALSE)
library(tinytex, quietly = TRUE, warn.conflicts = FALSE)
library(stringr, quietly = TRUE, warn.conflicts = FALSE)
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)
library(reshape2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(gtools, quietly = TRUE, warn.conflicts = FALSE)
library(here, quietly = T, warn.conflicts = F)
library(boot, quietly = T, warn.conflicts = F)

@@

setwd(paste0(here::here(), "/datasets/"))

@@

ggplot(samp, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), color = "darkorange") +
   geom_vline(xintercept = mu, col = "black", height = .8, lty = 2) +
   geom_vline(xintercept = mu - se, col = "darkred", lty = 2) +
   geom_vline(xintercept = mu + se, col = "darkred", lty = 2)

@@

```{r}
boot.fn <- function(data, index){
   mean(data[index]$values)
}

@@

cst.boot <- function(values, n) {
   bootstrap <- numeric(I)

@@

   alpha <- 0.05

@@

   list(observed = observed,
         mean = boot.mean,
         bias = boot.bias,
         se = boot.se,
         conf = quantile(bootstrap, c(alpha/2, 1 - alpha/2)))
}

@@

```{r}
n.50 <- cst.boot(samp$values, 50)

@@

```{r}
n.10 <- cst.boot(samp$values, 10)

@@

@@

@@

@@

@@

@@

>>>>

library(ggplot2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(ggthemes, quietly = TRUE, warn.conflicts = FALSE)
library(knitr, quietly = TRUE, warn.conflicts = FALSE)
library(kableExtra, quietly = TRUE, warn.conflicts = FALSE)
library(Rblpapi, quietly = TRUE, warn.conflicts = FALSE)
library(scales, quietly = TRUE, warn.conflicts = FALSE)
library(pander, quietly = TRUE, warn.conflicts = FALSE)
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(formattable, quietly = TRUE, warn.conflicts = FALSE)
library(grid, quietly = TRUE, warn.conflicts = FALSE)
library(gridExtra, quietly = TRUE, warn.conflicts = FALSE)
library(png, quietly = TRUE, warn.conflicts = FALSE)
library(extrafont, quietly = TRUE, warn.conflicts = FALSE)
library(tinytex, quietly = TRUE, warn.conflicts = FALSE)
library(stringr, quietly = TRUE, warn.conflicts = FALSE)
library(lubridate, quietly = TRUE, warn.conflicts = FALSE)
library(reshape2, quietly = TRUE, warn.conflicts = FALSE)
library(ggrepel, quietly = TRUE, warn.conflicts = FALSE)
library(gtools, quietly = TRUE, warn.conflicts = FALSE)
library(here, quietly = T, warn.conflicts = F)
library(boot, quietly = T, warn.conflicts = F)
library(ggExtra, quietly = T, warn.conflicts = F)

@@

data.dir <- paste0(here::here(), "/datasets/")
```

@@

ggplot(samp, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), color = "darkorange") +
   geom_vline(xintercept = mu, col = "black", lty = 2) +
   geom_vline(xintercept = mu - se, col = "darkred", lty = 2) +
   geom_vline(xintercept = mu + se, col = "darkred", lty = 2)

@@

boot.fn <- function(data, index){
   mean(data[index]$values)
}

@@

cst.boot <- function(values, n, I = 10e3, alpha = 0.05) {
   bootstrap <- numeric(I)

@@

@@

   list(bootstrap = bootstrap,
         observed = observed,
         mean = boot.mean,
         bias = boot.bias,
         se = boot.se,
         conf = quantile(bootstrap, c(alpha/2, 1 - alpha/2)))
}

@@

```{r}
n <- 50
samp <- data.table(values = rnorm(n, mean = mu, sd = sigma))
n.50 <- cst.boot(samp$values, n)

@@

```{r}
n <- 10
samp <- data.table(values = rnorm(n, mean = mu, sd = sigma))
n.10 <- cst.boot(samp$values, 10)

@@

n <- 200; r <- 5; lambda <- 4; mu <- lambda/r

pop <- data.table(values = rgamma(n, shape = lambda, rate = r))

ggplot(pop, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")
```

@@

```{r}
samp <- data.table(values = sample(pop$values, n, replace = T))

ggplot(samp, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")

xbar <- mean(samp$values); sd <- sd(samp$values)

xbar; sd
```

@@

```{r}
I <- 10e3

boot.fn <- function(data, index) {
   mean(data[index]$values)
}

boot(samp, boot.fn, R = I)

n.200 <- cst.boot(samp$values, n)

ggplot(data.table(values = n.200$bootstrap), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange") +
   labs(title = "N=200, bootstrapped mean (Gamma r=5, lambda = 4)")
```

@@

```{r}
tbl <- data.table(Data = c("Population", "Sampling Distribution", "Sample", "Bootstrap Distribution"),
                  Mean = c(mu, sd(samp$values), n.200$observed, n.200$mean),
                  SD = c(mu, mu/sqrt(n), sd(samp$values), n.200$se))

@@

```{r}
n <- 50

pop <- data.table(values = rgamma(n, shape = lambda, rate = r))

ggplot(pop, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")

samp <- data.table(values = sample(pop$values, n, replace = T))

ggplot(samp, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")

xbar <- mean(samp$values); sd <- sd(samp$values)

xbar; sd

n.50 <- cst.boot(samp$values, n)

ggplot(data.table(values = n.50$bootstrap), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange") +
   labs(title = "N=50, bootstrapped mean (Gamma r=5, lambda = 4)")

tbl <- data.table(Data = c("Population", "Sampling Distribution", "Sample", "Bootstrap Distribution"),
                  Mean = c(mu, sd(samp$values), n.50$observed, n.50$mean),
                  SD = c(mu, mu/sqrt(n), sd(samp$values), n.50$se))

```{r}
n <- 10

pop <- data.table(values = rgamma(n, shape = lambda, rate = r))

ggplot(pop, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")

samp <- data.table(values = sample(pop$values, n, replace = T))

ggplot(samp, aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")

xbar <- mean(samp$values); sd <- sd(samp$values)

xbar; sd

n.50 <- cst.boot(samp$values, n)

ggplot(data.table(values = n.50$bootstrap), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange") +
   labs(title = "N=50, bootstrapped mean (Gamma r=5, lambda = 4)")

tbl <- data.table(Data = c("Population", "Sampling Distribution", "Sample", "Bootstrap Distribution"),
                  Mean = c(mu, sd(samp$values), n.50$observed, n.50$mean),
                  SD = c(mu, mu/sqrt(n), sd(samp$values), n.50$se))

@@

```{r}
ne <- 14 # n even
no <- 15 # n odd

wwe <- rnorm(ne) # draw samples of size ne
wwo <- rnorm(no) # draw random samples of size no

N <- 10^4

even.boot <- numeric(N) # save space
odd.boot <- numeric(N)

for(i in 1:N)
{
   x.even <- sample(wwe, ne, replace = T)
   x.odd <- sample(wwo, no, replace = T)

   even.boot[i] <- median(x.even)
   odd.boot[i] <- median(x.odd)
}

boot <- data.table(even = even.boot, odd = odd.boot)

p1 <- ggplot(boot, aes(even)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange") +
   labs(title = "even")

p2 <- ggplot(boot, aes(odd)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange") +
   labs(title = "odd")

gridExtra::grid.arrange(p1, p2)
```

Change the sample sizes to 36 and 37; 200 and 201; and 10,000 and 10,001.

Note the similarities/dissimalarities, trends, and so on. Why does the parity of the sample size matter? (_Note: Adjust the x limits in the plots as needed._)

```{r}
ne <- 36 # n even
no <- 37 # n odd

wwe <- rnorm(ne) # draw samples of size ne
wwo <- rnorm(no) # draw random samples of size no

N <- 10^4

even.boot <- numeric(N) # save space
odd.boot <- numeric(N)

for(i in 1:N)
{
   x.even <- sample(wwe, ne, replace = T)
   x.odd <- sample(wwo, no, replace = T)

   even.boot[i] <- median(x.even)
   odd.boot[i] <- median(x.odd)
}

boot <- data.table(even = even.boot, odd = odd.boot)

p1 <- ggplot(boot, aes(even)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

p2 <- ggplot(boot, aes(odd)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

gridExtra::grid.arrange(p1, p2)
```

```{r}
ne <- 200 # n even
no <- 201 # n odd

wwe <- rnorm(ne) # draw samples of size ne
wwo <- rnorm(no) # draw random samples of size no

N <- 10^4

even.boot <- numeric(N) # save space
odd.boot <- numeric(N)

for(i in 1:N)
{
   x.even <- sample(wwe, ne, replace = T)
   x.odd <- sample(wwo, no, replace = T)

   even.boot[i] <- median(x.even)
   odd.boot[i] <- median(x.odd)
}

boot <- data.table(even = even.boot, odd = odd.boot)

p1 <- ggplot(boot, aes(even)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

p2 <- ggplot(boot, aes(odd)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

gridExtra::grid.arrange(p1, p2)
```

```{r}
ne <- 10000 # n even
no <- 10001 # n odd

wwe <- rnorm(ne) # draw samples of size ne
wwo <- rnorm(no) # draw random samples of size no

N <- 10^4

even.boot <- numeric(N) # save space
odd.boot <- numeric(N)

for(i in 1:N)
{
   x.even <- sample(wwe, ne, replace = T)
   x.odd <- sample(wwo, no, replace = T)

   even.boot[i] <- median(x.even)
   odd.boot[i] <- median(x.odd)
}

boot <- data.table(even = even.boot, odd = odd.boot)

p1 <- ggplot(boot, aes(even)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

p2 <- ggplot(boot, aes(odd)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(even), col = "darkorange")

gridExtra::grid.arrange(p1, p2)
```

For odd n, median will be one of the sample points. For smaller n, there will be only n possible values for the median, so the sampling distribution is more "granular" than when n is even.

### 5.11

Import the data from data set Bangladesh. In addition to aresnic concentrations for 271 wells, the data set contains cobolt and chlorine concentrations.

a.) Conduct EDA on the chlorine concentrations and describe the salient features.

```{r}
Bangladesh <- data.table(read.csv(paste0(data.dir, "Bangladesh.csv"),
                                 header = T))

head(Bangladesh)

ggplot(Bangladesh, aes(Chlorine)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(Chlorine), col = "darkorange")

GGally::ggpairs(Bangladesh)
```

b.) Bootstrap the mean.

```{r}
N <- 10e3

boot.fn <- function(data, index){
   mean(data[index], na.rm = T)
}

boot(Bangladesh$Chlorine, boot.fn, R = N)

observed <- mean(Bangladesh$Chlorine, na.rm = T)

bootstrap <- numeric(N)

for(i in 1:N)
{
   bootstrap[i] <- mean(sample(Bangladesh$Chlorine, size = nrow(Bangladesh), replace = T), na.rm = T)   
}

ggplot(data.table(values = bootstrap), aes(values)) +
   geom_histogram(aes(y = ..density.., fill = ..count..), bins = 30) +
   geom_density(aes(values), col = "darkorange")
```

c.) Find and interpret the 95% bootstrap percentile confidence interval.

```{r}
alpha <- 0.05
quantile(bootstrap, c(alpha/2, 1 - alpha/2))
```

The spread on the confidence interval is extremely large, which is unsuprising given the heavly skewed distribtion of the sample.

d.) What is the bootstrap estimate of the bias? What fraction of the bootstrap standard error does it represent?

```{r}
bias <- observed - mean(bootstrap)

sd(bootstrap) / bias
```
